---
title: "Statistical Analysis in R - York Footfall Data Analysis Report"
author: "Joao Marques Hassun"
date: "2024-10-19"
output: html_document
---

---

## Section 1

# Load the libraries.

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(gridExtra)
library(kableExtra)

```

This report fulfills the request of determining which is the best location to place stalls in order to gauge daily footfall.

1. Load in the data and check it for integrity/quality.

2. Provide a summary table that shows the following for each location where footfall was measured: 
  - The date of the first and last day when footfall was measured at this location 
  - The mean daily footfall 
  - The standard deviation of the daily footfall 
  - The highest daily footfall 
  - The lowest daily footfall

3. Plot the distribution of footfall across all days in the dataset. This should show the distributions separated by the locations where footfall is measured.

4. Perform a t-test to compare the footfall on Coney Street to Stonegate.

5. Perform an additional t-test to compare the footfall on Coney Street to Stonegate only on weekends.

---

## Data Dictionary

This data was captured by cameras that automatically measure the number of people who walk past a certain location: the 'footfall'. The variables are described in the table below:

Variable | Description
------------- | -------------
Date| The data in which the data was recorded
SiteName | The name of the site in which the data was captured, which is York for all.
LocationName| The specific location in which the data was recorded
WeekDay| The day of the week in which the data was recorded
TotalCount| The total count of 'Footfall' recorded 
Recording_ID| The unique identifier for people leaving footfall
Footfall| The measure of the number of people who walk past a certain location

---

# Read the csv file and assign it to a variable.

```{r}

Footfall <- read_csv("York_Footfall_data.csv")

```

# Plot the data to check the distribution.

```{r, warning=FALSE, message=FALSE}

FootfallScatterPlot <- ggplot(data = Footfall) + geom_point(mapping = aes(x = Date, y = TotalCount, color = LocationName)) + labs(title = "Distribution Scatter Plot for Footfall", x = "Date", y = "Total Count", color = "Location Name") + theme(plot.title = element_text(hjust = 0.5))

FootfallScatterPlot

```

# Create a summary of the data to check its integrity.

```{r}

summary(Footfall)

# The summary reveals that the SiteName, LocationName and WeekDay columns all have the expected amount of values.

# Creating a summary enables the identification of NA values in the TotalCount and Recording_ID columns.

# To check that the remaining columns also have the expected amount of values (8204), the length function is used.

length(Footfall$Date)

length(Footfall$TotalCount)

length(Footfall$Recording_ID)

```

# Data cleaning.

```{r}
# Considering how far apart min and max values are for the TotalCount column, we replace NAs with the median. 

# Since Recording_ID is an identification measure to each individual leaving footfall, it is not necessary to replace or remove its NAs.

Footfall$TotalCount[is.na(Footfall$TotalCount)] <- median(Footfall$TotalCount, na.rm = TRUE)

# Display Footfall's summary again to verify that the NAs present in Total Count were replaced with the median.

summary(Footfall)

```

# Create a summary table which displays information for each location. 

```{r}

Summary_Footfall <- Footfall %>%
  group_by(LocationName) %>%
  summarise(
    FirstDay = min(Date),
    LastDay = max(Date),
    MeanFootfall = round(mean(TotalCount), 0),
    StandardDeviationFootfall = round(sd(TotalCount), 0),
    MaxFootfall = max(TotalCount),
    MinFootfall = min(TotalCount)
  ) %>%
  knitr::kable(col.names = c("Location Name", "First Date", "Last Date", "Mean Footfall", "Standard Deviation of Footfall", "Max Footfall", "Min Footfall"))

# Since footfall is a discrete measure, the round function is used to transform continuous values into discrete values.

Summary_Footfall

```

# Create a new dataframe with only the 2019 information.

```{r}

max(Footfall$Date) 

# This function displays the last date in which footfall was measured, confirming the data does not extend post 2019.

Data_2019 <- Footfall %>% 
  filter(grepl(2019, Date)) 

Data_2019

```

# Plot the distribution of footfall across all days in the dataset.

```{r}

ScatterPlotFootfall2019 <- ggplot(data=Data_2019) + geom_point(mapping=aes(x=TotalCount, y=Date, color = LocationName)) + facet_wrap(.~LocationName) + labs(title = "Scatter Plot for Footfall in 2019", x = "Total Count", y = "Date", color = "Location Name") + theme(plot.title = element_text(hjust = 0.5))

ScatterPlotFootfall2019

WeekDayScatterPlot2019 <- ggplot(data=Data_2019) + geom_point(mapping=aes(x=WeekDay, y=TotalCount, color = LocationName)) + facet_grid(LocationName~.) + labs(title = "Footfall Scatter Plot by Days in 2019", x = "Week Day", y = "Total Count", color = "Location Name") + theme(plot.title = element_text(hjust = 0.5))

WeekDayScatterPlot2019

```

# Create a smaller dataframe with only the information regarding Coney Street and Stonegate to make the t-test process easier.

```{r}

ConeyStreet.vs.Stonegate <- Data_2019 %>%
  filter(LocationName %in% c("Coney Street", "Stonegate")) %>%
  select(LocationName, WeekDay, TotalCount)

ConeyStreet.vs.Stonegate

```

# Perform a t-test to compare the footfall on Coney Street to Stonegate.

```{r}

CvS_t_test_result <- t.test(TotalCount ~ LocationName, data = ConeyStreet.vs.Stonegate)

CvS_t_test_result

```

# Create a smaller dataframe with only the information regarding Coney Street and Stonegate on weekends to make the process for the additional t-test easier.

```{r}

ConeyStreet.vs.Stonegate.on.Weekends <- Data_2019 %>%
  filter(LocationName %in% c("Coney Street", "Stonegate"),
         WeekDay %in% c("Saturday", "Sunday")) %>%
  select(LocationName, WeekDay, TotalCount)

ConeyStreet.vs.Stonegate.on.Weekends

```

# Perform an additional t-test to compare the footfall on Coney Street to Stonegate on weekends.

```{r}

CvS_Weekend_t_test <- t.test(TotalCount ~ LocationName, data = ConeyStreet.vs.Stonegate.on.Weekends)

CvS_Weekend_t_test

```

## Section 2

This report presents the analysis conducted to determine the most optimal location for placing promotional stalls in York based on daily footfall data. The workflow included loading the relevant R libraries and the York Footfall dataset, validating data integrity, generating summary statistics, and conducting statistical tests.

The dataset includes pedestrian traffic counts collected across six locations: Church Street, Coney Street, Micklegate, Parliament Street, Parliament Street at M&S, and Stonegate. Each record includes Date, SiteName, LocationName, WeekDay, TotalCount, and Recording_ID. Initial exploration revealed a few outliers in Stonegate during December 2017; however, since the analysis focuses solely on data from 2019, these were not excluded.

Missing values were observed in the TotalCount and Recording_ID columns. To retain as much data as possible, TotalCount NAs were replaced with the median (16,990), considering the wide range between the minimum (402) and maximum (328,310). Since Recording_ID serves as a unique identifier, its NAs were left unchanged.

Visual exploration of 2019 data indicated that Coney Street, Stonegate, and Parliament Street had the highest daily footfall counts, while Micklegate had the lowest. A t-test comparing daily footfall between Coney Street and Stonegate across the full week revealed that Coney Street had a significantly higher mean (M = 20,817.45) than Stonegate (M = 19,204.09);
t(699.18) = 3.36, p < .05.
This result suggests a statistically significant difference, supporting the recommendation of Coney Street for weekday stall placement.

A second t-test, restricted to weekends, found that Stonegate's mean (M = 26,166.96) was slightly higher than Coney Street's (M = 25,863.37);
t(203.88) = -0.29, p = .772.
However, the result is not statistically significant, indicating no clear difference in weekend footfall between the two locations.